Embedding Intuitionistic into Classical Logic

### 2022-10-13 FoSSaCS/fossacs-EIICL.tex
Symmetric difference

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic in order to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We transfer there methods to the first order case and also give an embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

@@

\begin{wrapfigure}[6]{r}{0.4\textwidth}
		\vspace*{-1.7cm}
		\begin{tabular}{l|c|c}
			Embedding&571&21.4\%\\
			 Embedding (idem.)&557&20.9\%\\
			ileanCoP 1.2&875&32.8\%\\
			nanoCoP-i 2.0&858&32.1\%\\\hline
			Total&2669&100\%
		\end{tabular}
		\caption{Solved problems.}
		\label{fig:a}
\end{wrapfigure}

We have benchmarked our first-order translation on version 1.1.2 of the ILTP problem set~\cite{iltp} by performing our translation and then examining the generated classical problem using Vampire~\cite{kovacs2013first}. For comparison the state-of-the-art intuitionistic provers ileanCoP 1.2~\cite{otten2008leancop} and nanoCoP-i 2.0~\cite{otten2021nanocop} are used. All benchmarks were conducted on an Intel Core i5-8400 2.80GHz machine running Linux with a time limit of 10s per problem. Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results can be seen in Figure \ref{fig:a}. In general our approach falls short of the state of the art. However when binning for $|\mathcal Q|$ and $|\mathcal X|$ as seen in Figure~\ref{fig:b} we observe something interesting.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.75\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3-5	&16&		&17&16.2\%		&50&47.6\%		&40&28.1\%	&105\\
		0&6-10	&16&		&4&21.1\%		&9&47.4\%		&7&36.8\%	&19\\
		0&11-20	&2&			&0&0\%			&2&5.9\%		&3&8.8\%	&34\\
		0&21+	&8&			&7&41.2\%		&9&52.9\%		&1&5.9\%	&17\\
		1-2&0	&14&		&14&82.4\%		&16&94.1\%		&15&88.2\%	&17\\
		1-2&1-2	&70&		&67&24.8\%		&211&78.1\%		&206&76.3\%	&270\\
		1-2&3-5	&80&		&77&24.7\%		&141&45.2\%		&122&39.1\%	&312\\
		1-2&6-10&16&		&15&23.4\%		&29&45.3\%		&28&43.8\%	&64\\
		1-2&11+	&6&			&6&16.2\%		&10&27.0\%		&9&24.3\%	&37\\
		3-5&0	&14&		&14&87.5\%		&16&100\%		&16&100\%	&16\\
		3-5&1-2	&14&		&11&22.4\%		&37&75.5\%		&36&73.4\%	&49\\
		3-5&3-5	&10&		&15&9.4\%		&72&45.3\%		&58&36.5\%	&159\\
		3-5&6-10&18&		&16&8.0\%		&64&32.2\%		&83&41.7\%	&199\\
		3-5&11+	&2&			&2&5.3\%		&0&0\%			&3&7.9\%	&38\\
		6+&0-2	&1&			&1&2.0\%		&49&98.0\%		&50&100\%	&50\\
		6+&3+	&7&			&3&	0.6\%		&24&3.4\%		&24&3.4\%	&605
	\end{tabular}
	\caption{Solved problems binned by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}
Leveraging the power of Vampire, for low $|\mathcal X|$ and $|\mathcal Q|$ the embedding performs on par or even outperforms the dedicated intuitionistic provers. For larger values the state of the art provers have an advantage, in particular when both parameters are combined. Therefore our tool is already useful if it is known that $|\mathcal X|$ and $|\mathcal Q|$ are small. We see that idempotence does not improve performance. On the other hand from a model-theoretic perspective this is quite powerful. We suspect that a more clever encoding of this property could improve our results. There are other restriction on models we could consider encoding based on structural properties of the formula, e.g. the relation between occurrences of different atoms.
Another interesting observation is that $|\mathcal X|$ and $|\mathcal Q|$ also seem to correlate with success rate for the dedicated intuitionistic provers.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\usepackage{tikz}
\usetikzlibrary{shapes,backgrounds}

\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic in order to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We transfer there methods to the first order case and also give an embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

\begin{wrapfigure}[8]{r}{0.4\textwidth}
	\vspace*{-.5cm}
	\begin{tabular}{l|c|c}
		Embedding&571&21.4\%\\
		Embedding (idem.)&557&20.9\%\\
		ileanCoP 1.2&875&32.8\%\\
		nanoCoP-i 2.0&858&32.1\%\\\hline
		Total&2669&100\%
	\end{tabular}
	\caption{Solved problems.}
	\label{fig:a}
\end{wrapfigure}

@@

\begin{wrapfigure}[13]{r}{0.4\textwidth}
	\vspace*{4.9cm}
	\hspace*{1cm}
	\begin{tikzpicture}
		\begin{scope}[opacity=0.3, blend mode=multiply]
			\def\firstcircle{(90:.75) circle (1.5)}
			\def\secondcircle{(210:.75) circle (1.5)}
			\def\thirdcircle{(330:.75) circle (1.5)}
			\begin{tikzpicture}
				\begin{scope}[shift={(3cm,-5cm)}, fill opacity=0.5]
					\fill[red] \firstcircle;
					\fill[green] \secondcircle;
					\fill[blue] \thirdcircle;
					\draw node {$491$};
					\draw node at (30:1.2) {$30$};
					\draw node at (150:1.2) {$42$};
					\draw node at (270:1.2) {$22$};
					\draw node at (90:1.8) {$8$};
					\draw node at (210:1.8) {$2$};
					\draw node at (330:1.8) {$417$};
				\end{scope}
			\end{tikzpicture}
		\end{scope}
	\end{tikzpicture}
	\vspace*{-7.1cm}
	\caption{Symmetric difference. Instances proven by Embedding (red), Embedding (idem.) (green), ileanCoP or nanoCoP (blue)}
	\label{fig:c}
\end{wrapfigure}

We have benchmarked our first-order translation on version 1.1.2 of the ILTP problem set~\cite{iltp} by performing our translation and then examining the generated classical problem using Vampire~\cite{kovacs2013first}. For comparison the state-of-the-art intuitionistic provers ileanCoP 1.2~\cite{otten2008leancop} and nanoCoP-i 2.0~\cite{otten2021nanocop} were used. All benchmarks were conducted on an Intel Core i5-8400 2.80GHz machine running Linux with a time limit of 10s per problem. The problem \textsc{SYN007+1} was removed as our normal form procedure would explode the size of instance.

Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results can be seen in Figure \ref{fig:a}. In general our approach falls short of the state of the art. However when binning for $|\mathcal Q|$ and $|\mathcal X|$ as seen in Figure~\ref{fig:b} we observe good performance for $|\mathcal Q| = 0$ and $|\mathcal X| \leq 3$. Figure~\ref{fig:c} shows that we solve 52 problems not solved by the existing provers.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.8\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3-5	&16&15.2\%	&17&16.2\%		&50&47.6\%		&40&28.1\%	&105\\
		0&6+	&27&38.6\%	&11&15.7\%		&20&28.6\%		&11&15.7\%	&70\\
		1-2&0-2	&84&29.3\%	&81&28.2\%		&227&79.1\%		&221&77.0\%	&287\\
		1-2&3-5	&80&25.6\%	&77&24.7\%		&141&45.2\%		&122&39.1\%	&312\\
		1-2&6+	&22&21.8\%	&21&20.8\%		&39&38.6\%		&37&36.6\%	&101\\
		3-5&0-5	&38&17.0\%	&40&17.9\%		&125&55.8\%		&110&49.1\%	&224\\
		3-5&6+  &20&8.4\%	&18&7.6\%		&64&27.0\%		&86&36.3\%	&237\\
		6+&0+	&8&1.1\%	&5&0.7\%		&73&9.7\%		&74&9.8\%	&755
	\end{tabular}
	\caption{Solved problems binned by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}

 Therefore our tool is already useful if it is known that $|\mathcal X|$ and $|\mathcal Q|$ are small. We see that idempotence does not improve performance. On the other hand from a model-theoretic perspective it is quite powerful. We suspect that a more clever encoding of this property could improve our results. There are other restrictions on models we could consider, based on structural properties of the formula, e.g. the relation between occurrences of different atoms.
